Collaborative Multi-output Gaussian Processes

### 2014-06-16 paper/inference.tex
final edits

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
In variational inference, we find the ``closest'' approximate distribution to the true posterior in terms of the $\KL$ divergence.
We first observe that the true posterior distribution can be written as:
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of  inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used by \citet{titsias2009variational} and \citet{hensmangaussian} 
to derive variational inference algorithms for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \eqref{eq:gu} and \eqref{eq:hv}), 
we only need to 
learn $q(\u, \v)$ so as to minimize the divergence between the approximate posterior and
the true posteriors.
The quality of approximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.

@@

Learning of the inducing inputs, which was not considered in the single output case in \citet{hensmangaussian}, is also possible in our stochastic optimization approach.
In the supplementary material, we show that 
%with the help of vectorized operations, 
the additional cost of computing the derivatives of the lower bound wrt the inducing inputs is not significantly 
higher than the cost of updating the variational parameters.
%can be computed with a cost \emph{independent of the input dimensionality}.
This makes optimizing the inducing locations a practical option, which can be critical 
in high-dimensional problems. %where fixing the inducing inputs may hinder the performance of the model.
Indeed, our experiments on a large scale multi-output problem show that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation. 
%
\subsection{COMPLEXITY ANALYSIS \label{sec:complexity}}
In this section we analyze the complexity of the model and compare it to existing multi-output approaches.
For consistency, we first unify common notations used for all models.
We use $P$ as the number of outputs, $N$ as the number of inputs, $Q$ as the number of shared latent processes, $M_t$ as the \emph{total} number of inducing inputs.
It is worth noting that $M_t = (P + Q) \times M$ in our COGP model, assuming that each sparse process has equal number of inducing points. 
Also, COGP have $P$ additional individual processes, one for each output.

>>>>

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
In variational inference, we find the ``closest'' approximate distribution to the true posterior in terms of the $\KL$ divergence.
We first observe that the true posterior distribution can be written as:
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of  inducing variables is the sufficient statistics of the corresponding latent process. 
This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used by \citet{titsias2009variational} and \citet{hensmangaussian} 
to derive variational inference algorithms for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \eqref{eq:gu} and \eqref{eq:hv}), 
we only need to 
learn $q(\u, \v)$ so as to minimize the divergence between the approximate posterior and
the true posteriors.
The quality of approximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.

@@

Learning of the inducing inputs, which was not considered in the single output case in \citet{hensmangaussian}, is also possible in our stochastic optimization approach.
In the supplementary material, we show that 
%with the help of vectorized operations, 
the additional cost of computing the derivatives of the lower bound wrt the inducing inputs is not significantly 
higher than the cost of updating the variational parameters.
%can be computed with a cost \emph{independent of the input dimensionality}.
This makes optimizing the inducing locations a practical option, which can be critical 
in high-dimensional problems. %where fixing the inducing inputs may hinder the performance of the model.
Indeed, our experiments on a large scale multi-output problem show that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation. 
%
\subsection{COMPLEXITY ANALYSIS \label{sec:complexity}}
In this section we analyze the complexity of the model and compare it to existing multi-output approaches.
For consistency, we first unify common notations used for all models.
We use $P$ as the number of outputs; $N$ as the number of inputs; $Q$ as the number of shared latent processes; and $M_t$ as the \emph{total} number of inducing inputs.
It is worth noting that $M_t = (P + Q) \times M$ in our COGP model, assuming that each sparse process has equal number of inducing points. 
Also, COGP has $P$ additional individual processes, one for each output.

### 2014-06-16 paper/intro.tex
final edits

% Gaussian processes are good for bayesian regression, Very flexible in multi-output settings, for e.g. bonilla, alvarez, wilson
Gaussian process models  \citep[GPs,][]{rasmussen-williams-book}  are a popular choice in 
Bayesian regression due to their ability to capture complex dependencies and non-linearities in data.
In particular, when having multiple outputs or tasks they have proved effective in modeling 
the dependencies between the tasks  and  they have been shown to outperform competitive baselines and 
single output learners  \citep{bonilla-et-al-nips-08,teh-et-al-aistats-05,alvarez-lawrence-nips-08,wilson-et-al-icml-12}.
% However they are costly
However, the prohibitive cost of performing exact inference in  GP models severely hinders 
their application to  large scale multi-output problems. For example, na\"{i}ve inference 
in a fully coupled Gaussian process model over $P$  outputs and $N$ data-points can have 
a complexity of $\calO(N^3P^3)$  and $\calO(N^2P^2)$ in time and memory, respectively.

@@

% Tranditional approaches in single output setting, based e..g. on inducing points are not good enough 
In the single output setting previous attempts to scale up  GP inference 
resort to approximate inference. Most  approximation techniques 
 can be understood 
within a single probabilistic framework that uses a set of \emph{inducing points}  in order to 
obtain an approximate process (or a low-rank approximate covariance) over which inference can be 
performed more efficiently \citep{quinonero2005unifying}. These models have been referred 
to in the literature as sparse models. 
Nevertheless, straightforward application of such approximate techniques will yield a computational 
cost of at least $\calO(PNM^2)$ in time and  $\calO(PNM)$ in memory, where 
$M$ is the number of inducing points. This high complexity still prevents us from applying GP models 
to  large scale multi-output problems.

% Useful observations
In this work we approach the challenge of building scalable multi-output Gaussian process models  
based on the following observations.
Firstly, \emph{inducing variables} are the key catalyst
for achieving sparsity and dealing with large scale problems in Gaussian process models.
In particular, they capture the \emph{sufficient statistics} of a dataset allowing the construction of sparse 
processes that can approximate arbitrarily well the exact GP model \citep{titsias2009variational}.
Secondly, the use of global latent variables (such as the inducing points) allows us to 
induce dependencies in a highly correlated model efficiently. 
This observation is exploited in \cite{hensmangaussian} for single output GP regression models 
where, by explicitly representing a distribution over the inducing variables, stochastic variational inference 
can be used to work with millions of data points.
Finally, the key to multi-output and multi-task learning is to model dependencies between the outputs  
based on realistic assumptions of what can be shared across the tasks. It turns out that 
sharing  ``sparsity structure'' can not only be a reasonable assumption but also a crucial component
when modeling dependencies between different related tasks.

% Our approach starts from mixing latent process
% having inducing points for sparsity
% sharing inducing points across the latent process
Based on these observations, we propose the collaborative multi-output Gaussian Process (COGP) model 
where each output is a weighted combination of sparse latent processes.
 Each latent process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
Correlations among the outputs are fostered via sharing of multiple inducing point sets 
which in turn create a medium through which joint inference happens efficiently.
% exploit stochastic variational inference
In particular, we derive a variational lower bound of the model evidence that decomposes across 
inputs and outputs. 
This decomposition makes possible the application of stochastic variational inference, 
thus allowing the model to handle a large number of observations per output and a large number of outputs.
Furthermore, learning of all  the  parameters in the model, including
kernel hyperparameters and inducing inputs, can be done in a single stochastic optimization framework.

>>>>

% Gaussian processes are good for bayesian regression, Very flexible in multi-output settings, for e.g. bonilla, alvarez, wilson
Gaussian process models  \citep[GPs,][]{rasmussen-williams-book}  are a popular choice in 
Bayesian regression due to their ability to capture complex dependencies and non-linearities in data.
In particular, when having multiple outputs or tasks they have proved effective in modeling 
the dependencies between the tasks, outperforming  competitive baselines and 
single output learners  \citep{bonilla-et-al-nips-08,teh-et-al-aistats-05,alvarez-lawrence-nips-08,wilson-et-al-icml-12}.
% However they are costly
However, the prohibitive cost of performing exact inference in  GP models severely hinders 
their application to  large scale multi-output problems. For example, na\"{i}ve inference 
in a fully coupled Gaussian process model over $P$  outputs and $N$ data points can have 
a complexity of $\calO(N^3P^3)$  and $\calO(N^2P^2)$ in time and memory, respectively.

@@

% Tranditional approaches in single output setting, based e..g. on inducing points are not good enough 
In the single output setting previous attempts to scale up  GP inference 
resort to approximate inference. Most  approximation methods  
 can be understood 
within a single probabilistic framework that uses a set of \emph{inducing points}  in order to 
obtain an approximate process (or a low-rank approximate covariance) over which inference can be 
performed more efficiently \citep{quinonero2005unifying}. These models have been referred 
to in the literature as sparse models. 
Nevertheless, straightforward application of such approximate techniques will yield a computational 
cost of at least $\calO(PNM^2)$ in time and  $\calO(PNM)$ in memory, where 
$M$ is the number of inducing points. This high complexity still prevents us from applying GP models 
to  large scale multi-output problems.

% Useful observations
In this work we approach the challenge of building scalable multi-output Gaussian process models  
based on the following observations.
Firstly, inducing variables are the key catalyst
for achieving sparsity and dealing with large scale problems in Gaussian process models.
In particular, they capture the \emph{sufficient statistics} of a dataset allowing the construction of sparse 
processes that can approximate arbitrarily well the exact GP model \citep{titsias2009variational}.
Secondly, the use of global latent variables (such as the inducing points) allows us to 
induce dependencies in a highly correlated model efficiently. 
This observation is exploited in \cite{hensmangaussian} for single output GP regression models 
where, by explicitly representing a distribution over the inducing variables, stochastic variational inference 
can be used to work with millions of data points.
Finally, the key to multi-output and multi-task learning is to model dependencies between the outputs  
based on realistic assumptions of what can be shared across the tasks. It turns out that 
sharing  ``sparsity structure'' can not only be a reasonable assumption but also a crucial component
when modeling dependencies between different related tasks.

% Our approach starts from mixing latent process
% having inducing points for sparsity
% sharing inducing points across the latent process
Based on these observations, we propose the collaborative multi-output Gaussian Process (COGP) model 
where latent processes are mixed to generate dependent outputs. 
Each process is sparse and characterized by its own set of inducing points.
The sparsity structure enabling output correlations is thus created via the shared inducing sets. 
To learn this structure, we maintain an explicit representation of the posterior over the inducing points which in turn allows inference to be carried out efficiently.
% exploit stochastic variational inference
In particular, we obtain a variational lower bound of the model evidence that decomposes across 
inputs and outputs. 
This decomposition makes possible the application of stochastic variational inference, 
thus allowing the model to handle a large number of observations per output and a large number of outputs.
Furthermore, learning of all  the  parameters in the model, including
kernel hyperparameters and inducing inputs, can be done in a single stochastic optimization framework.

### 2014-06-16 paper/model.tex
final edits

Equations \eqref{eq:u} and \eqref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (Equations  \eqref{eq:gu} and \eqref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the \emph{sufficient statistics} role of $\u$ and $\v$ in the model.
Here by sufficient statistics we mean, for any sparse process (say $g_j$), any other set of function values is 
independent of $\g_j$ given the inducing variables $\u_j$.
%
\subsection{LIKELIHOOD MODEL}
As mentioned above, we assume that  observations for each output 
are  (noisy) linear combinations of the $Q$ latent functions $g_{j}(\vec{x})$ plus
an independent function $h_{i}(\vec{x})$. Hence we have that the likelihood 
with standard iid Gaussian noise is given by:
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P \prod_{n=1}^N \Normal( y_{in} ; \sum_{j=1}^Q w_{ij} g_j(\x_n) + h_i(\x_n), \beta_i^{-1}),
\end{align}
where $w_{ij}$ are the corresponding weights and $\beta_i$ is the precision of each Gaussian.
%which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor model \citep{teh-et-al-aistats-05}.
However, doing so is against the purpose of our model which encourages sharing of outputs via the inducing variables.
Furthermore, as we shall see in the next section, explicit representation of these variables is fundamental to scalable inference of the model.

>>>>

Equations \eqref{eq:u} and \eqref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (Equations  \eqref{eq:gu} and \eqref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the sufficient statistics role of $\u$ and $\v$ in the model.
Here by sufficient statistics we mean, for any sparse process (say $g_j$), any other set of function values is 
independent of $\g_j$ given the inducing variables $\u_j$.
%
\subsection{LIKELIHOOD MODEL}
As mentioned above, we assume that  observations for each output 
are  (noisy) linear combinations of the $Q$ latent functions $g_{j}(\vec{x})$ plus
an independent function $h_{i}(\vec{x})$. Hence we have that the likelihood 
with standard iid Gaussian noise is given by:
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P \prod_{n=1}^N \Normal( y_{in} ; \sum_{j=1}^Q w_{ij} g_j(\x_n) + h_i(\x_n), \beta_i^{-1}),
\end{align}
where $w_{ij}$ are the corresponding weights and $\beta_i$ is the precision of each Gaussian.
%which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor model \citep{teh-et-al-aistats-05}.
However, doing so is against the purpose of our model which encourages sharing of outputs via the inducing variables.
Furthermore, as we shall see in the next section, explicit representation of these variables is fundamental to scalable inference of the model.
